High Performance P$^{3}$M N-body code: CUBEP$^3$M

### 2012-08-20 MNRAS_V1/paper_new.tex
Ready to Commit?

For instance, a measurement of the baryonic acoustic oscillation (BAO) dilation 
scale can provide tight constraints on the dark energy equation of scale 
\citep{Eisenstein:2005su,2006PhRvD..74l3507T, 2007MNRAS.381.1053P,2009arXiv0902.4680S}. 
The most optimal estimates of the uncertainty requires the 
knowledge of the matter power spectrum covariance matrix, which is only accurate 
when measured from a large sample of N-body simulations \citep{2005MNRAS.360L..82R, 
2009ApJ...700..479T, 2011ApJ...726....7T}. For the same reasons, the most accurate 
estimates of weak gravitational lensing signal is obtained by propagating photons 
in past light cones that are extracted from simulated density fields 
\citep{2003ApJ...592..699V, 2009ApJ...701..945S, 2009A&A...499...31H}.
Another area where large-scale N-body simulations have in recent years been 
instrumental are in simulations of early cosmic structures and reionization 
\citep[e.g.][]{2006MNRAS.369.1625I,2007ApJ...654...12Z,2007ApJ...671....1T,
2011arXiv1107.4772I}. The reionization process is primarily driven by low-mass 
galaxies, which for both observational and theoretical reasons, need to be resolved 
in fairly large volumes, which demands simulations with a very large dynamic range.   
%{\bf (Ilian, could you say something about reionization here? Just general references as to why one needs N-body codes to perform 
%accurate predictions/forecasts, etc.)}

@@

With the advent of large computing facilities, parallel computations have now become 
common practice, and N-body codes have evolved both in performance and complexity. 
Many have opted for  `tree' algorithms, including {\small GADGET} \citep{2001NewA....6...79S, 2005MNRAS.364.1105S}, {\small PMT} \citep{1995ApJS...98..355X}, {\small GOTPM} \citep{2004NewA....9..111D}, and Hydra \citep{1995ApJ...452..797C}, in which the local resolution increases with the density of the matter field. 
These often have the advantage to balance the work load across the computing units, which enable fast calculations even in high density regions. 
The drawback is a significant loss in speed, which can be only partly recovered by turning off the tree algorithm. 
The same reasoning applies to mesh-refined codes  \citep{1991ApJ...368L..23C}, 
which in the end are not designed to perform fast PM calculations on large scales.

@@

 The standard package also contains support for gas cosmological evolution through a portable TVD-MHD module \citep{2003ApJS..149..447P} that scales up to thousands of cores as well (see \citet{2010arXiv1004.1680P} and footnote 4 in section \ref{sec:scaling}), and a coupling interface with the radiative transfer code C2-Ray \citep{2006NewA...11..374M}.
 {\small CUBEP$^3$M} is therefore one of the most competitive and versatile public N-body code, 
 and has been involved in a number of scientific applications over the last few years,
spanning the field of weak lensing \citep{Vafaei10, 2008MNRAS.388.1819L,  2009arXiv0905.0501D, 2010PhRvD..81l3015L, 2010arXiv1012.0444Y, 2012arXiv1202.2332H},  BAO
 \citep{2010arXiv1008.3506Z,  2012MNRAS.419.2949N, 2012MNRAS.423.2288H, 2012arXiv1205.4989H}, 
formation of early cosmic structures \citep{2008arXiv0806.2887I,2010arXiv1005.2502I},
observations of dark stars \citep{2010MNRAS.407L..74Z,2012MNRAS.tmp.2794I} and
reionization \citep{2011arXiv1107.4772I,Fernandez:2011ab,2011MNRAS.413.1353F,2012MNRAS.422..926M,Datta:2011hv,2012arXiv1203.0517F}. 
Continuous efforts are being made to develop, extend and improve the code and each of its utilities, 
and we expect that this will pave the way to an increasing number of science projects. 
Notably, the fact that the force calculation is multi-layered makes the code extendible, 
and opens the possibility to run {\small CUBEP$^3$M} on hybrid CPU-GPU clusters.
It is thus important for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code.

@@

Because the coarse grid arrays require $4^3$ times less memory per node, 
it does not contribute much to the total memory requirement, and the bulk of the foot-print is 
concentrated in a handful of fine grid arrays.
Some of these are only required for intermediate steps of the calculations, 
hence it is possible to hide therein many coarse grid arrays\footnote{ This memory recycling is done with `equivalence' statements in {\small F90}}.   
We present here the largest arrays used by the code:
\begin{enumerate}
\item{{\tt xv} stores the position and velocity of each particle} 
\item{{\tt ll} stores the linked-list that accelerate the access to particles in each coarse grid cell}
\item{{\tt send\_buf, recv\_buf} store the particles to be {\small MPI}-passed to the neighbouring sub-volumes}
\item{{\tt rho\_f} and {\tt cmplx\_rho\_f} store 
the local fine grid density  in real and Fourier space respectively}
\item{{\tt force\_f} stores the force of gravity (short range only) along the three Cartesian directions}
\item{{\tt kern\_f} stores the fine grid force kernel in the three directions}
\item{{\tt PID} stores the unique particle identification tags.}
\end{enumerate}
The particle ID is a feature that can be switched off by removing a compilation flag, 
and allows to optimize the code for higher resolution configurations.

{\bf In terms of memory, {\small CUBEP$^3$M} can be configured such as the main footprint is dominated exclusively by the 
particle phase space array, the linked list and the buffer exchange arrays, in which case it asymptotes to about $44$ bytes per particles. 
For a fixed problem size, this can be achieved by maximizing the number of tiles, thereby reducing the size of the local fine mesh arrays.
There is a limit at which we can do such a volume breakdown, though, since the physical 
volume on each tile must be at least twice as large as the buffer, i.e. $\ge 96^3$ cells (see section \ref{sec:Poisson}).
In addition, this allows a high rate of particle exchange across nodes;  the memory per particle can be brought down to $32$ bytes 
if little particle passing is expected. 
However, minimizing the memory does not optimize the speed, as many {\small CPU}-intensive operations are performed on each tile -- {\small FFTW} for instance.
A higher number of cells per tile with fewer tiles usually run faster, in which case the other arrays in the above-mentioned list become memory heavy.
%
%uses about $130 - 150$ bytes per particles, and is in that sense more heavy than other codes like the 
For comparison, Lean-{\small GADGET-2} code uses  $84$ per particles \citep{Gadget2Lean}.}

@@

Since these calculations are performed on two grids of different resolution, a sampling window function must be convoluted 
both with the density and the kernel (see [Eq. 7-8] of MPT).
When matching the two force kernels, the long range force is always on the low side close to the cutoff region, whereas the short range force is uniformly scattered across the theoretical $1/r^2$ value -- intrinsic features of the CIC and NGP interpolation schemes respectively.  By performing force measurements on two particles randomly placed in the volume, we identified a small region surrounding the cutoff length in which we empirically adjust both kernels such as to improve the match. Namely, for $14 \le r \le 16$, $\mbox{\boldmath $\omega$}_{s}({r}) \rightarrow 0.985\mbox{\boldmath $\omega$}_{s}({ r})$,
and for  $12 \le r \le 16$, $\mbox{\boldmath $\omega$}_{l}({r}) \rightarrow 1.2\mbox{\boldmath $\omega$}_{l}({ r})$.
{\bf A recent adjustment of the similar kind was also applied on the fine kernel to compensate for a systematic underestimate of the force at grid distances.
This is caused by the uneven balance of the force about the $1/r^2$ law in the NGP interpolation scheme.
Although most figures in this paper were made prior this correction, the new code configuration applies a 90 per cent boost to 
the kernel for the six elements that are adjacent, i.e. $\mbox{\boldmath $\omega$}_{s}({r=1}) \rightarrow 1.9\mbox{\boldmath $\omega$}_{s}({r=1})$ (see section \ref{subsec:force} for more details).}

@@

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/power_rsoft.eps}
  \caption{Dark matter power spectrum, measured at $z=0.5$ in a series of SciNet256 simulations in which the softening length is varied.
  The simulations started off at $z= 100$, and have a box size of $100 h^{-1}\mbox{Mpc}$.
    \label{fig:rsoft}}
\end{center}
\end{figure}

@@

\begin{figure*}%[ht]
%  \vskip -0.5cm 
  \begin{center}
    \includegraphics[width=3.0in]{graphs/scaling_cubep3m_curie.eps}
    \includegraphics[width=3.0in]{graphs/scaling_cubep3m_new.eps}
%  \vskip -1.2cm 
%  \vskip -0.5cm 
  \caption{Scaling of {\small CUBEP$^3$M} on Curie fat nodes (left) and 
    on Ranger TACC facility for very large number of cores (right). Plotted is the code speedup 
    ($N_{\rm particles}^3/t_{\rm wallclock}$) against core count, normalized by the smallest run 
    in each case. Dashed line indicates the ideal weak 
    scaling. The data are listed in Table \ref{summary_scaling_table}.
    \label{scaling}
% \vskip -0.9cm 
}
\end{center}
\end{figure*}

@@

\begin{figure*}%[ht]
  \begin{center}
%    \includegraphics[width=5.2in]{graphs/0.000xz_w_halos_pink_blue_enh-1.eps}
\includegraphics[width=5.2in]{graphs/0.000proj_1000x1000.eps}
  \caption{Detail of a dark matter density $3.2 h^{-1}\mbox{Gpc}$Êper side, projected over $15 h^{-1}\mbox{Mpc}$ and measured at $z=0$ in a RANGER4000 simulation.
  Particles are shown in pink, haloes in blue in the online version, and the colour scale is linear. The full projection is 64 times larger,
  but we show here only a sub-volume, which provides more details on the structure.
    \label{fig:density}}
\end{center}
\end{figure*}

\begin{figure*}%[ht]
  \begin{center}
    \includegraphics[width=5.2in]{graphs/power_highres.eps}
  \caption{Dark matter power spectrum, measured at $z=0$ in a RANGER4000 simulation,
  compared to the non-linear predictions of {\small HALOFIT}. 
  The vertical line corresponds to the scale of the coarse mesh, while the dotted line represents the Poisson noise.
  The particle masses are of $5.68 \times 10^{10} M_{\odot}$. 
    \label{fig:power_highres}}
\end{center}
\end{figure*}

@@

\begin{figure*}%[ht]
  \begin{center}
 %   \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_new-1.eps}
   % \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_new.eps}
    \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_new_kern.eps}
    \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_new_kern.eps}
    %\includegraphics[width=3.0in]{graphs/densityForce_ppext=0_rebin_new-1.eps}
   \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_rebin_new_kern.eps}
   \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_rebin_new_kern.eps}
 %   \includegraphics[width=3.0in]{graphs/DeltaF_pp0.eps}
   \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_rebin_new_kern_N10.eps}
   \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_rebin_new_kern_N10.eps}
%    \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_rebin_N10_new-1.eps}
   % \includegraphics[width=3.0in]{graphs/DeltaF_pp0_N10.eps}
%    \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_rebin_N10_new-1.eps}
  \caption{({\it top left}:) Gravity force in the P$^3$M algorithm, versus distance, compared with the exact $1/r^{2}$ law.
   Distance is in fine mesh cell units, and force in simulation units.
    This particular calculation was obtained in a CITA128 realization with  a box size of $100 h^{-1}\mbox{Mpc}$,
    in a single time step. 
    ({\it top right}:) Fractional error on the force in the P$^3$M algorithm in the radial direction (top) and fractional tangential contribution (bottom).
  In a full simulation run, the scatter averages over many time steps, 
    thanks to the inclusion of a random offset that is imposed on each particle, as discussed in the main text.
    ({\it middle row}:) Top row organized in 50 logarithmic bins; for the tangential contribution,  we plot only the scatter about the mean.  
    ({\it bottom row}:) Same as middle row, but averaged over ten time steps.
    \label{fig:den_force_fracErr}}
\end{center}
\end{figure*}

@@

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/power_w_wo_disp_mesh.eps}
  \caption{Dark matter power spectrum, measured at $z=180$, $100$, $20$ and $10$, in a CITA256 realization that is $1000 h^{-1}$ Mpc per side. 
  The dashed line represent the initial condition power spectrum, the dotted lines are the linear predictions, and  the open circles the standard P$^3$M configuration. 
  The dots were obtained by simply removing the random offset that is normally applied at each time step. \label{fig:disp_mesh}}
\end{center}
\end{figure}

@@

{\bf 
As mentioned in section \ref{sec:Poisson}, most of the tests presented in this paper were obtained without the fine force kernel correction at  $r=1$.
For completeness, we present here the motivation behind the recent improvement that compensates for a systematic underestimate
at grid distances. 
Without the correction, the force of gravity looks like Fig. \ref{fig:den_force_fracErr_kern}, 
where we observe that the fractional error in the radial direction is systematically on the low side, by up to 50 per cent. 
This is caused by the fact that in the strong regime of a $1/r^2$ law, the mean force from a cell of uniform density, as felt by a particle, 
is not equivalent to that of the same cell collapsed to its centre -- the actual position of equivalence would be at a closer distance.
Basically, the idea is to boost the small distance elements of the kernel by a small amount,
which thereby corrects for the loss of force above-described, and observed in Fig. \ref{fig:den_force_fracErr_kern}.

We  show in Fig. \ref{fig:fudge} the impact on the power spectrum of three different trials: a) the force from the first layer of neighbours is boosted by $90$ per cent, b)
the force from the first two layers of neighbours is boosted by $60$ per cent, and c) by $80$ per cent.
We observe that with the trial a), we are able to improve the power at turn around by a factor of $1.5$, which almost doubles the resolution.  
This test was performed in a series of SciNet256 simulations with a volume of $150 h^{-1} \mbox{Mpc}$ and at $z=0$.
We see from the figure that in the original configuration, the departure from the non-linear predictions occur at $k\sim 3.0 h\mbox{Mpc}^{-1}$,
whereas the first trial allows for a similar resolution down to $k\sim 4.5 h\mbox{Mpc}^{-1}$. 
The two other correction schemes are more aggressive, and although they recover power at even smaller scales, 
they exhibit a small overestimate compare to non-linear predictions in the range $2.0 < k < 4.0 h\mbox{Mpc}^{-1}$, 
and more testing should be done prior to using these corrections. 
We have also tried to increase even more the boost factor in the case a), but once again, the effect started to propagate to larger scales, 
an effect we need to avoid. %We show in Fig. \ref{fig:den_force_fracErr_kern} the impact of the trial a) on the force of gravity. 
%We observe that when averaged over ten time steps, the scatter  is five to ten times larger than with the original kernel, 
%both in the radial and tangential directions, but the radial part is better balanced about the theoretical prediction.

We therefore conclude that the fine force kernel should be subject to a boost factor following `trial a', i.e. 
$\mbox{\boldmath $\omega$}_{s}({r=1}) \rightarrow 1.9\mbox{\boldmath $\omega$}_{s}({r=1})$.
It is thinkable that other corrections could outperform this one, hence the kernel correction factor
should be considered as an adjustable parameter;
`no correction' is considered as the conservative mode, which is accurate only up to the turn around 
observed in the dimensionless power spectrum, and which has been used in most of the tests presented in this paper. 
Since higher resolution can be achieved at no cost with the corrected fine force kernel, it has now been adopted in the default configuration;
`trial b' and `trial c' should be considered as too aggressive. We note, again, that we have not exhausted the list of possible correction,
and that future work in that direction might result in even better resolution gain.}

@@

\begin{figure}%[ht]
  \begin{center}
%    \includegraphics[width=3.0in]{graphs/DeltaF_pp0_new_kern.eps}
  %  \includegraphics[width=3.0in]{graphs/DeltaF_pp0_new_kern_N10.eps}
  % \includegraphics[width=3.0in]{graphs/DeltaF_pp0_new_kern.eps}
    \includegraphics[width=3.0in]{graphs/DeltaF_pp0_N10.eps}
  \caption{Fractional error in the gravity force, without the fine force kernel correction at $r=1$.
      This particular calculation was obtained in a CITA128 realization with  a box size of $150 h^{-1}\mbox{Mpc}$,
     averaged over ten time steps.
    \label{fig:den_force_fracErr_kern}}
\end{center}
\end{figure}

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/power_fudge.eps}
  \caption{Dark matter power spectrum, measured at $z=0$, in a SciNet256 realization that is $150 h^{-1}$ Mpc per side. 
  The thick solid line represents predictions from {\small CAMB}, the thin solid line is the default (i.e. without correction) configuration of the code, and the three other lines represent
  three different corrections that are applied at grid distances to compensate for the systematic force underestimation seen in the bottom panels of Fig. \ref{fig:den_force_fracErr}. \label{fig:fudge}}
\end{center}
\end{figure}

@@

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/power_ra_max.eps}
  \caption{Dark matter power spectrum, measured at $z=0$ in a series of CITA256 realizations. 
 The starting redshift was raised to $z=200$ to enhance the systematic effect. The different curves show different values of $r_{max}$. 
  The resources required to run these simulations increase rapidly as $r_{max}$ decreases, as seen in Table \ref{table:ra_max}.    \label{fig:ra_max}}
\end{center}
\end{figure}

>>>>

For instance, a measurement of the baryonic acoustic oscillation (BAO) dilation 
scale can provide tight constraints on the dark energy equation of scale 
\citep{Eisenstein:2005su,2006PhRvD..74l3507T, 2007MNRAS.381.1053P,2009arXiv0902.4680S}. 
The most optimal estimates of the uncertainty requires the 
knowledge of the matter power spectrum covariance matrix, which is only accurate 
when measured from a large sample of N-body simulations \citep{2005MNRAS.360L..82R, 
2009ApJ...700..479T, 2011ApJ...726....7T}. For the same reasons, the most accurate 
estimates of weak gravitational lensing signal is obtained by propagating photons 
in past light cones that are extracted from simulated density fields 
\citep{2003ApJ...592..699V, 2009ApJ...701..945S, 2009A&A...499...31H}.
Another area where large-scale N-body simulations have in recent years been 
instrumental are in simulations of early cosmic structures and reionization 
\citep[e.g.][]{2006MNRAS.369.1625I,2007ApJ...654...12Z,2007ApJ...671....1T,
2012MNRAS.423.2222I}. The reionization process is primarily driven by low-mass 
galaxies, which for both observational and theoretical reasons, need to be resolved 
in fairly large volumes, which demands simulations with a very large dynamic range.   
%{\bf (Ilian, could you say something about reionization here? Just general references as to why one needs N-body codes to perform 
%accurate predictions/forecasts, etc.)}

@@

With the advent of large computing facilities, parallel computations have now become 
common practice, and N-body codes have evolved both in performance and complexity. 
Many have opted for  `tree' algorithms, including {\small GADGET} \citep{2001NewA....6...79S, 2005MNRAS.364.1105S}, {\small PMT} \citep{1995ApJS...98..355X}, {\small GOTPM} \citep{2004NewA....9..111D}, or adaptive P$^3$M codes like Hydra \citep{1995ApJ...452..797C}, in which the local resolution increases with the density of the matter field. 
These often have the advantage to balance the work load across the computing units, which enable fast calculations even in high density regions. 
The drawback is a significant loss in speed, which can be only partly recovered by turning off the tree algorithm. 
The same reasoning applies to mesh-refined codes  \citep{1991ApJ...368L..23C}, 
which in the end are not designed to perform fast PM calculations on large scales.

@@

 The standard package also contains support for gas cosmological evolution through a portable TVD-MHD module \citep{2003ApJS..149..447P} that scales up to thousands of cores as well (see \citet{2010arXiv1004.1680P} and footnote 4 in section \ref{sec:scaling}), and a coupling interface with the radiative transfer code C$^2$-ray \citep{2006NewA...11..374M}.
 {\small CUBEP$^3$M} is therefore one of the most competitive and versatile public N-body code, 
 and has been involved in a number of scientific applications over the last few years,
spanning the field of weak lensing \citep{Vafaei10, 2008MNRAS.388.1819L,  2009arXiv0905.0501D, 2010PhRvD..81l3015L, 2012MNRAS.421..832Y, 2012arXiv1202.2332H},  BAO
 \citep{2011ApJ...728...35Z,  2012MNRAS.419.2949N, 2012MNRAS.423.2288H, 2012arXiv1205.4989H}, 
formation of early cosmic structures \citep{2008arXiv0806.2887I,2010arXiv1005.2502I},
observations of dark stars \citep{2010MNRAS.407L..74Z,2012MNRAS.422.2164I} and
reionization \citep{2012MNRAS.423.2222I,2012ApJ...750...20F,2011MNRAS.413.1353F,2012MNRAS.422..926M,2012MNRAS.424.1877D,2012MNRAS.424..762D}. 
Continuous efforts are being made to develop, extend and improve the code and each of its utilities, 
and we expect that this will pave the way to an increasing number of science projects. 
Notably, the fact that the force calculation is multi-layered makes the code extendible, 
and opens the possibility to run {\small CUBEP$^3$M} on hybrid CP2010arXiv1012.0444YU-GPU clusters.
It is thus important for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code.

@@

Because the coarse grid arrays require $4^3$ times less memory per node, 
it does not contribute much to the total memory requirement, and the bulk of the foot-print is 
concentrated in a handful of fine grid arrays.
Some of these are only required for intermediate steps of the calculations, 
hence it is possible to hide therein many coarse grid arrays\footnote{ This memory recycling is done with `equivalence' statements in {\small F90}}.   
We present here the largest arrays used by the code:
\begin{enumerate}
\item{{\tt xv} stores the position and velocity of each particle} 
\item{{\tt ll} stores the linked-list that accelerate the access to particles in each coarse grid cell}
\item{{\tt send\_buf, recv\_buf} store the particles to be {\small MPI}-passed to the neighbouring sub-volumes}
\item{{\tt rho\_f} and {\tt cmplx\_rho\_f} store 
the local fine grid density  in real and Fourier space respectively}
\item{{\tt force\_f} stores the force of gravity (short range only) along the three Cartesian directions}
\item{{\tt kern\_f} stores the fine grid force kernel in the three directions}
\item{{\tt PID} stores the unique particle identification tags as double integers.}
\end{enumerate}
The particle ID is a feature that can be switched off by removing a compilation flag, 
and allows to optimize the code for higher resolution configurations.

 In terms of memory, {\small CUBEP$^3$M} can be configured such as the main footprint is dominated exclusively by the 
particle phase space array, the linked list and the buffer exchange arrays, in which case it asymptotes to about $44$ bytes per particles. 
For a fixed problem size, this can be achieved by maximizing the number of tiles, thereby reducing the size of the local fine mesh arrays.
There is a limit at which we can do such a volume breakdown, though, since the physical 
volume on each tile must be at least twice as large as the buffer, i.e. $\ge 96^3$ cells (see section \ref{sec:Poisson}).
In addition, this allows a high rate of particle exchange across nodes;  the memory per particle can be brought down to $32$ bytes 
if little particle passing is expected. 
However, minimizing the memory does not optimize the speed, as many {\small CPU}-intensive operations are performed on each tile -- {\small FFTW} for instance.
A higher number of cells per tile -- typically $176^3$ -- with fewer tiles usually run faster, in which case the other arrays in the above-mentioned list become memory heavy,
and the memory per particle is closer to 100 - 140 bytes. Also, enabling the particle ID adds about 24 bytes per particle: 8 for the main array, and 8 for the two {\small MPI}-passing buffers as well.
%uses about $130 - 150$ bytes per particles, and is in that sense more heavy than other codes like the 
For comparison, Lean-{\small GADGET-2} code uses  $84$ bytes per particles \citep{Gadget2Lean}.

@@

Since these calculations are performed on two grids of different resolution, a sampling window function must be convoluted 
both with the density and the kernel (see [Eq. 7-8] of MPT).
When matching the two force kernels, the long range force is always on the low side close to the cutoff region, whereas the short range force is uniformly scattered across the theoretical $1/r^2$ value -- intrinsic features of the CIC and NGP interpolation schemes respectively.  By performing force measurements on two particles randomly placed in the volume, we identified a small region surrounding the cutoff length in which we empirically adjust both kernels such as to improve the match. Namely, for $14 \le r \le 16$, $\mbox{\boldmath $\omega$}_{s}({r}) \rightarrow 0.985\mbox{\boldmath $\omega$}_{s}({ r})$,
and for  $12 \le r \le 16$, $\mbox{\boldmath $\omega$}_{l}({r}) \rightarrow 1.2\mbox{\boldmath $\omega$}_{l}({ r})$.
A recent adjustment of the similar kind was also applied on the fine kernel to compensate for a systematic underestimate of the force at grid distances.
This is caused by the uneven balance of the force about the $1/r^2$ law in the NGP interpolation scheme.
Although most figures in this paper were made prior this correction, the new code configuration applies a 90 per cent boost to 
the kernel for the six elements that are adjacent, i.e. $\mbox{\boldmath $\omega$}_{s}({r=1}) \rightarrow 1.9\mbox{\boldmath $\omega$}_{s}({r=1})$ (see section \ref{subsec:force} for more details).

@@

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/fig3.eps}
  \caption{Dark matter power spectrum, measured at $z=0.5$ in a series of SciNet256 simulations in which the softening length is varied.
  The simulations started off at $z= 100$, and have a box size of $100 h^{-1}\mbox{Mpc}$.
    \label{fig:rsoft}}
\end{center}
\end{figure}

@@

\begin{figure*}%[ht]
%  \vskip -0.5cm 
  \begin{center}
    \includegraphics[width=3.0in]{graphs/fig4a.eps}
    \includegraphics[width=3.0in]{graphs/fig4b.eps}
%  \vskip -1.2cm 
%  \vskip -0.5cm 
  \caption{Scaling of {\small CUBEP$^3$M} on Curie fat nodes (left) and 
    on Ranger TACC facility for very large number of cores (right). Plotted is the code speedup 
    ($N_{\rm particles}^3/t_{\rm wallclock}$) against core count, normalized by the smallest run 
    in each case. Dashed line indicates the ideal weak 
    scaling. The data are listed in Table \ref{summary_scaling_table}.
    \label{scaling}
% \vskip -0.9cm 
}
\end{center}
\end{figure*}

@@

\begin{figure*}%[ht]
  \begin{center}
%    \includegraphics[width=5.2in]{graphs/0.000xz_w_halos_pink_blue_enh-1.eps}
\includegraphics[width=5.2in]{graphs/fig5.eps}
  \caption{Detail of a dark matter density $3.2 h^{-1}\mbox{Gpc}$Ê per side, projected over $15 h^{-1}\mbox{Mpc}$ and measured at $z=0$ in a RANGER4000 simulation.
  Particles are shown in pink, haloes in blue in the online version, and the colour scale is linear. The full projection is 64 times larger,
  but we show here only a sub-volume, which provides more details on the structure.
    \label{fig:density}}
\end{center}
\end{figure*}

\begin{figure*}%[ht]
  \begin{center}
    \includegraphics[width=5.2in]{graphs/fig6.eps}
  \caption{Dark matter power spectrum, measured at $z=0$ in a RANGER4000 simulation,
  compared to the non-linear predictions of {\small HALOFIT}. 
  The vertical line corresponds to the scale of the coarse mesh, while the dotted line represents the Poisson noise.
  The particle masses are of $5.68 \times 10^{10} M_{\odot}$. 
    \label{fig:power_highres}}
\end{center}
\end{figure*}

@@

\begin{figure*}%[ht]
  \begin{center}
 %   \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_new-1.eps}
   % \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_new.eps}
    \includegraphics[width=3.0in]{graphs/fig7a.eps}
    \includegraphics[width=3.0in]{graphs/fig7b.eps}
    %\includegraphics[width=3.0in]{graphs/densityForce_ppext=0_rebin_new-1.eps}
   \includegraphics[width=3.0in]{graphs/fig7c.eps}
   \includegraphics[width=3.0in]{graphs/fig7d.eps}
 %   \includegraphics[width=3.0in]{graphs/DeltaF_pp0.eps}
   \includegraphics[width=3.0in]{graphs/fig7e.eps}
   \includegraphics[width=3.0in]{graphs/fig7f.eps}
%    \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_rebin_N10_new-1.eps}
   % \includegraphics[width=3.0in]{graphs/DeltaF_pp0_N10.eps}
%    \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_rebin_N10_new-1.eps}
  \caption{({\it top left}:) Gravity force in the P$^3$M algorithm, versus distance, compared with the exact $1/r^{2}$ law.
   Distance is in fine mesh cell units, and force in simulation units.
    This particular calculation was obtained in a CITA128 realization with  a box size of $100 h^{-1}\mbox{Mpc}$,
    in a single time step. 
    ({\it top right}:) Fractional error on the force in the P$^3$M algorithm in the radial direction (top) and fractional tangential contribution (bottom).
  In a full simulation run, the scatter averages over many time steps, 
    thanks to the inclusion of a random offset that is imposed on each particle, as discussed in the main text.
    ({\it middle row}:) Top row organized in 50 logarithmic bins; for the tangential contribution,  we plot only the scatter about the mean.  
    ({\it bottom row}:) Same as middle row, but averaged over ten time steps.
    \label{fig:den_force_fracErr}}
\end{center}
\end{figure*}

@@

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/fig8.eps}
  \caption{Dark matter power spectrum, measured at $z=180$, $100$, $20$ and $10$, in a CITA256 realization that is $1000 h^{-1}$ Mpc per side. 
  The dashed line represent the initial condition power spectrum, the dotted lines are the linear predictions, and  the open circles the standard P$^3$M configuration. 
  The dots were obtained by simply removing the random offset that is normally applied at each time step. \label{fig:disp_mesh}}
\end{center}
\end{figure}

@@

As mentioned in section \ref{sec:Poisson}, most of the tests presented in this paper were obtained without the fine force kernel correction at  $r=1$.
For completeness, we present here the motivation behind the recent improvement that compensates for a systematic underestimate
at grid distances. 
Without the correction, the force of gravity looks like Fig. \ref{fig:den_force_fracErr_kern}, 
where we observe that the fractional error in the radial direction is systematically on the low side, by up to 50 per cent. 
This is caused by the fact that in the strong regime of a $1/r^2$ law, the mean force from a cell of uniform density, as felt by a particle, 
is not equivalent to that of the same cell collapsed to its centre -- the actual position of equivalence would be at a closer distance.
Basically, the idea is to boost the small distance elements of the kernel by a small amount,
which thereby corrects for the loss of force above-described, and observed in Fig. \ref{fig:den_force_fracErr_kern}.

We  show in Fig. \ref{fig:fudge} the impact on the power spectrum of three different trials: a) the force from the first layer of neighbours is boosted by $90$ per cent, b)
the force from the first two layers of neighbours is boosted by $60$ per cent, and c) the force from two layers is boosted by $80$ per cent.
We observe that with the trial a), we are able to improve the power at turn around by a factor of $1.5$, which almost doubles the resolution.  
This test was performed in a series of SciNet256 simulations with a box length of $150 h^{-1} \mbox{Mpc}$ and at $z=0$.
We see from the figure that in the original configuration, the departure from the non-linear predictions occur at $k\sim 3.0 h\mbox{Mpc}^{-1}$,
whereas the first trial allows for a similar resolution down to $k\sim 4.5 h\mbox{Mpc}^{-1}$. 
The two other correction schemes are more aggressive, and although they recover power at even smaller scales, 
they exhibit a small overestimate compare to non-linear predictions in the range $2.0 < k < 4.0 h\mbox{Mpc}^{-1}$, 
and more testing should be done prior to using these corrections. 
We have also tried to increase even more the boost factor in the case a), but once again, the effect started to propagate to larger scales, 
an effect we need to avoid. %We show in Fig. \ref{fig:den_force_fracErr_kern} the impact of the trial a) on the force of gravity. 
%We observe that when averaged over ten time steps, the scatter  is five to ten times larger than with the original kernel, 
%both in the radial and tangential directions, but the radial part is better balanced about the theoretical prediction.

We therefore conclude that the fine force kernel should be subject to a boost factor following `trial a', i.e. 
$\mbox{\boldmath $\omega$}_{s}({r=1}) \rightarrow 1.9\mbox{\boldmath $\omega$}_{s}({r=1})$.
It is thinkable that other corrections could outperform this one, hence the kernel correction factor
should be considered as an adjustable parameter;
`no correction' is considered as the conservative mode, which is accurate only up to the turn around 
observed in the dimensionless power spectrum, and which has been used in most of the tests presented in this paper. 
Since higher resolution can be achieved at no cost with the corrected fine force kernel, it has now been adopted in the default configuration;
`trial b' and `trial c' should be considered as too aggressive. We note, again, that we have not exhausted the list of possible correction,
and that future work in that direction might result in even better resolution gain.

@@

\begin{figure}%[ht]
  \begin{center}
%    \includegraphics[width=3.0in]{graphs/DeltaF_pp0_new_kern.eps}
  %  \includegraphics[width=3.0in]{graphs/DeltaF_pp0_new_kern_N10.eps}
  % \includegraphics[width=3.0in]{graphs/DeltaF_pp0_new_kern.eps}
    \includegraphics[width=3.0in]{graphs/fig9.eps}
  \caption{Fractional error in the gravity force, without the fine force kernel correction at $r=1$.
      This particular calculation was obtained in a CITA128 realization with  a box size of $150 h^{-1}\mbox{Mpc}$,
     averaged over ten time steps.
    \label{fig:den_force_fracErr_kern}}
\end{center}
\end{figure}

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/fig10.eps}
  \caption{Dark matter power spectrum, measured at $z=0$, in a SciNet256 realization that is $150 h^{-1}$ Mpc per side. 
  The thick solid line represents predictions from {\small CAMB}, the thin solid line is the default (i.e. without correction) configuration of the code, and the three other lines represent
  three different corrections that are applied at grid distances to compensate for the systematic force underestimation seen in the bottom panels of Fig. \ref{fig:den_force_fracErr}. \label{fig:fudge}}
\end{center}
\end{figure}

@@

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/fig11.eps}
  \caption{Dark matter power spectrum, measured at $z=0$ in a series of CITA256 realizations. 
 The starting redshift was raised to $z=200$ to enhance the systematic effect. The different curves show different values of $r_{max}$. 
  The resources required to run these simulations increase rapidly as $r_{max}$ decreases, as seen in Table \ref{table:ra_max}.    \label{fig:ra_max}}
\end{center}
\end{figure}
